\Large{\textbf{A biophysical minimal model to investigate age-related changes in CA1 pyramidal cell electrical activity}}

### 2024-07-11 paper/article.tex
Update manuscript files

\documentclass[12pt]{article}
\input{settings}

% bib options
\let\oldbibliography\thebibliography
\renewcommand{\thebibliography}[1]{\oldbibliography{#1}
\setlength{\itemsep}{0pt}} 
\usepackage[numbers,square,sort,compress]{natbib}

% for symbols
\usepackage{gensymb}
\usepackage{textcomp}

% landscape pages
\usepackage{lscape}

% for tables
\usepackage{booktabs}
\renewcommand{\arraystretch}{1.5}

% long tables that run over 1pg
\usepackage{longtable}
\usepackage{array}

@@

\allowdisplaybreaks
\usepackage[parfill]{parskip}

% margins and lines
\usepackage{titlesec}
\usepackage[margin=2.5cm]{geometry}
\pagestyle{fancy} 
\fancyheadoffset[R]{0.2cm}
\usepackage{lineno}

% editing and track changes
%\usepackage{ulem}
\newcommand{\edit}[1]{\textcolor{blue}{#1}}

% title and authors
\title{\Large{\textbf{A biophysical minimal model to investigate age-related changes in CA1 pyramidal cell electrical activity}}}
\author[1*]{Erin C. McKiernan} 
\author[2**]{Marco A. Herrera-Valdez}
\author[3,4]{Diano F. Marrone}

\affil[1]{\small{Departamento de F\'isica, Facultad de Ciencias, Universidad Nacional Aut\'onoma de M\'exico}}
\affil[2]{\small{Laboratorio de Din\'amica, Biof\'isica y Fisiolog\'ia de Sistemas, Departamento de Matem\'aticas, Facultad de Ciencias, Universidad Nacional Aut\'onoma de M\'exico}}
\affil[3]{\small{Department of Psychology, Wilfrid Laurier University}}
\affil[4]{\small{McKnight Brain Institute, University of Arizona}}
\affil[*]{\small{Corresponding author: emckiernan@ciencias.unam.mx}}
\affil[**]{\small{Corresponding author: marcoh@ciencias.unam.mx}}

\date{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

\setlength{\jot}{10pt}

\titlespacing\section{0pt}{13pt plus 3pt minus 2pt}{0pt plus 2pt minus 2pt}
\titlespacing\subsection{0pt}{13pt plus 5pt minus 2pt}{0pt plus 2pt minus 2pt}
\titlespacing\subsubsection{0pt}{13pt plus 5pt minus 2pt}{0pt plus 2pt minus 2pt}

% specify headers
\fancyhf{}
\lhead{McKiernan, Herrera-Valdez, \& Marrone}
\rhead{\thepage}

\begin{abstract}
Aging is a physiological process that is still poorly understood, especially with respect to effects on the brain. There are open questions about aging that are difficult to answer with an experimental approach. Underlying challenges include  the difficulty of recording \textit{in vivo} single cell and network activity simultaneously with submillisecond  resolution, and brain compensatory mechanisms triggered by genetic, pharmacologic, or behavioral manipulations. Mathematical modeling can help address some of these questions by allowing us to fix parameters that cannot be controlled experimentally and investigate neural activity under different conditions. We present a biophysical minimal model of CA1 pyramidal cells (PCs) based on general expressions for transmembrane ion transport derived from basic thermodynamical principles. The model allows directly varying the contribution of ion channels by changing their number. By analyzing the dynamics of the model, we find parameter ranges that reproduce the variability in electrical activity seen in PCs. In addition, increasing the L-type {\Ca} channel expression in the model reproduces age-related changes in excitability that are qualitatively and quantitatively similar to those observed in PCs from aged animals.
\end{abstract}

\vspace{2cm}

\begin{minipage}{0.2\textwidth}
\includegraphics[width=0.8\textwidth]{figures/ccby.png}
\end{minipage}%
\begin{minipage}{0.75\textwidth}
\begin{scriptsize}
This work is licensed under a \href{https://creativecommons.org/licenses/by/4.0/}{Creative Commons Attribution 4.0 International License}, which allows users to copy, redistribute, remix, transform, and build upon the material. Please attribute the authors and cite this work as specified by the hosting preprint server and using its persistent identifier (DOI).
\end{scriptsize}
\end{minipage}

\newpage

@@

\section{Introduction}
As we age, our brains undergo many changes
\citep{oh2010learning,rosenzweig2003impact}, but we understand relatively little about these and their effects on neural function. What does
normal neurophysiological aging look like and what are the various stages? How does the electrical activity of neurons change and what are the biophysics underlying those changes?  How do
aging neurons respond to input from other cells? Answering these questions is not
just fundamental to understanding aging as a neurophysiological process, but also to understanding how this process may be altered in age-related disorders of clinical importance such as Alzheimer's \cite{fjell2014normal} and Parkinson's \cite{rodriguez2015parkinson} disease.

Many aging studies have focused on the hippocampus, an area of the brain involved in learning, memory formation, and spatial processing
\citep{oh2010learning,rosenzweig2003impact}. Aged rats \citep{barnes1985age,gage1984spatial,caprioli1991spatial} and humans \citep{wilkniss1997age} show impaired learning of hippocampal-dependent spatial tasks. Long-term potentiation (LTP), a proposed physiological substrate of memory formation, has been investigated in the hippocampus and its induction and maintenance
shown to be impaired in aged rats
\citep{deupree1993age,rosenzweig1997role}. A short-term form of plasticity, frequency potentiation/facilitation
(FP/FF), is also impaired in hippocampal pyramidal cells (PCs) from aged rats and correlates
with learning deficits
\citep{landfield1978impaired}.

Plasticity changes and behavioral impairments may result in part from altered {\Ca} signaling
in aged neurons
\citep{oh2010learning,rosenzweig2003impact}. Compared to CA1 PCs from young animals, PCs from aged animals show larger and longer post-burst afterhyperpolarizations (AHPs)
\citep{landfield1984prolonged,gant2009action,power2002age}. AHPs
are mediated by {\Ca}-dependent {\K} currents, which can act like brakes on the electrical activity of CA1 PCs
\citep{alger1980epileptiform,hotson1980calcium}. As a result, PCs show increased spike frequency adaptation and fire fewer action potentials (APs) in response to acute stimuli or during bursting activity 
\citep{gant2006early,moyer1992nimodipine,tombaugh2005slow}. Larger AHPs are associated with increased intracellular {\Ca}, mediated in part
by {\Ca} entry via L-type channels
\citep{campbell1996aging,moyer1992nimodipine,power2002age,thibault2001elevated}. Aged
animals show increases in L-type channel expression and/or channel
density at the plasma membrane
\citep{herman1998up,thibault1996increase,veng2002regionally,nunez2014surface}.
Animals with higher {\Ca} channel density perform poorly in spatial
tasks 
\citep{thibault1996increase}, while blockers of L-type channels can
restore learning and plasticity in older animals
\citep{norris1998reversal,sandin1990aging}.

It is not well understood how changes
in ion
channel gene expression and hippocampal PC excitability may affect neuron
responsiveness and microcircuit output. In part, this is due
to challenges inherent in performing the needed experiments. Single PCs are difficult
to access in intact animals where hippocampal microcircuit function is
preserved. It is also difficult to tease apart the influence of the many different neurophysiological factors that
change during aging. 
Mathematical modeling provides means to understand more about the effects of aging on hippocampal cellular excitability by controlling factors we cannot control experimentally.

Our previous work shows that mathematical expressions for different passive and active ion transport mechanisms can be derived from first principles of thermodynamics \cite{herrera2012membranes,herrera2013relating} using a common functional form \cite{herrera2018thermodynamic}. This results in a realistic representation of ionic flow across the membrane, and allows the model to reproduce phenomena such as rectification of ion currents seen in recordings. We present a model that reproduces the diversity of firing patterns in CA1 PC recordings, including adaptive firing, stimulus-induced bursting, and spontaneous bursting \cite{mckiernan2017ca1}. In addition, we reproduce several electrophysiological characteristics of aging by varying the expression of {\Ca} channels in the model, and make predictions about bursting activity in aged CA1 PCs, which to our knowledge has not been reported. We believe this model is ideal to further study the effects of various biophysical changes in CA1 PCs during aging, as well as potentially forming the basis for biophysical, yet computationally inexpensive, network models.

\section{Methods}

\subsection{Model}
To simulate the electrical activity of CA\edit{1} PCs, we used an extended version of a two-dimensional, biophysical model previously developed and characterized by two of the present authors  \cite{herrera2012membranes,herrera2013relating,herrera2018thermodynamic}. The equations for the ionic currents are derived from first principles of thermodynamics. Previous modeling studies have shown that to reproduce firing behaviors such as spike frequency adaptation and bursting, the minimum number of variables is three \cite{hindmarsh1984model,avron1993basic}. 
In particular, {\Ca} dynamics are important for producing adaptation and burst firing in CA1 PCs (for review see \cite{mckiernan2017ca1}). 
The model dynamics are therefore described by three ordinary differential equations for the time-dependent changes in the transmembrane potential ($v$, in mV), the proportion of open {\K} channels ($w$ in [0,1]), and the intracellular {\Ca} concentration ($c$, in $\mu$M), respectively   \citep{herrera2018thermodynamic}. Based on a well-known relationship between voltage-dependent activation of delayed rectifier {\K} channels and inactivation of {\Na} channels, $w$ also represents the proportion of inactivated {\Na} channels \cite{rinzel1985excitation,avron1991minimal}.

It is assumed that the membrane potential changes due to currents produced by ions transported across the membrane. We take into account currents mediated by a voltage-gated, inactivating {\Na} channel ($I_{NaT}$), voltage- and {\Ca}-gated \K channels ($I_{DK}$ and $I_{SK}$, respectively), and a {\Na}/{\K}-ATPase ($I_{NaK}$). The time-dependent change in membrane potential can be written as
\begin{equation}
C_m \partial_t v= I_{F} - I_{NaT}(v,w) - I_{CaL}(v,c) - I_{DK}(v,w) - I_{SK}(v,c) - I_{NaK}(v). \label{eq:dvdt}
\end{equation}
with $\partial_t$ representing the instantaneous change with respect to time. $C_m$ (pF) is a constant representing the change in the density of charge around the membrane with respect to voltage, typically referred to as membrane capacitance in models based on electrical circuits  \cite{herrera2020nonequivalent}. Based on recordings from rat CA1 PCs, it is  assumed that $C_m=25$  pF \cite{groc2002vivo}.

All the currents in Eq.~\eqref{eq:dvdt} are modeled using the same generic functional form, a product
\begin{equation*}
    I_x = a_x G_x    \varphi_x, 
\end{equation*}
 $x \in \lrSet{NaT,CaL,DK,SK,NaK}$, where $a_x$ is a whole-cell current amplitude (pA), $G_x$ is a gating term (between 0~and~1), and $\varphi_x$ is an adimensional term describing the driving force for the transmembrane flux (Table~\ref{tab:ionFluxes}). The terms $a_x= s_x N_x$, $x \in \lrSet{NaT,CaL,DK,SK,NaK}$ are whole-cell current amplitudes with
$s_x$ (pA) representing the current flowing through a single channel (or pump), and $N_x$ representing the number of membrane proteins mediating the current (e.g. {number of {\K} channels}). Of interest, $s_x$ is $\sim$1 pA for most voltage-gated channels \citep{hille2001ion}, and is $\sim$5-10 pA for SK channels \cite{stocker20042+}.

Assuming that none of the currents in the model exhibit rectification  \citep{herrera2018thermodynamic}, which agrees with recordings of the included currents, the adimensional component of the transmembrane flux can  be simplified and written as
\begin{equation}
    \varphi_x(v) = 2\eta_x \sinh\left(\eta_x \frac{v-v_x}{2v_T}\right),
\end{equation} 
where $\eta_x$ represents the number of charges transported across the membrane in a single transport event, and $v_x$ is the reversal potential for the current,   
 $x \in \left\{NaT,CaL,DK,SK,NaK \right\}$. For channels, $v_x$ is the Nernst potential for the ion \citep{herrera2018thermodynamic}. Of note,  $\eta_x=1$ for $x \in \lrSet{DK,SK,NaK}$ and $\eta_{NaT}=-1$   \citep{herrera2018thermodynamic}, which gives 
\begin{equation*}
    \varphi_x(v) = 2 \sinh\left( \frac{v-v_{x}}{2v_T}\right),
\end{equation*} 
for $x \in \lrSet{NaT,DK,SK,NaK}$.
For {\Ca} channels, the total charge transported by one ion crossing the membrane from the extracelular space is $\eta_{CaL}=-2$, so
\begin{equation*}
    \varphi_{Ca}(v) = 4 \sinh\left( \frac{v-v_{Ca}}{v_T}\right).
\end{equation*}

The driving force for flux is assumed to be the same for DK and SK channels. Therefore, the label $K$ is used for both fluxes from here on.   
The thermal potential $v_T=kT/q$ (mV), where $k$ is Boltzmann's constant (mJ/$^o$K), $T$ is the  absolute temperature ($^o$K), and $q$ is the elementary charge  (Coulombs). The Boltzmann constant can be thought of as a scaling factor between macroscopic (thermodynamic temperature) and microscopic (thermal energy) physics \citep{kalinin2005boltzmann}. The reversal potentials for the different currents depend on the  Nernst potentials for each ion, as given by 
\begin{equation}
v_x = \frac{v_T}{z_x} \textrm{ln} \left(\frac{[x]_o}{[x]_i} \right), \quad x \in  \left\{Na,Ca,K \right\}
\end{equation}
where $z_x$ is the ion valence and $[x]_o$ and $[x]_i$ are the ion concentrations outside and inside the cell, respectively. The reversal potential for the {\Na}/{\K}-ATPase is given by ${v_{NaK}=v_{ATP}+3v_{Na}-2v_{K}}$ \citep{herrera2018thermodynamic}. The Nernst potentials for {\Na} and {\K} are assumed to be constant, but $v_{Ca}$ varies because the intracellular {\Ca} concentration is a state variable in the model.

>>>>

% Template for PLoS
% Version 3.6 Aug 2022
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage[nopatch=eqnum]{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

% commands for ions and and math symbols
\newcommand{\Ca}{Ca$^{2+}$}
\newcommand{\K}{K$^{+}$}
\newcommand{\Na}{Na$^{+}$}
\newcommand{\lrRound}[1]{\left(#1\right)}
\newcommand{\lrSquare}[1]{\left[#1\right]}
\newcommand{\lrSet}[1]{\left\{#1\right\}}

% packages and commands for tables
\usepackage{booktabs}
\usepackage{longtable}
\renewcommand{\arraystretch}{1.5}

@@

% for symbols
\usepackage{gensymb}

% track changes in color
\newcommand{\edit}[1]{\textcolor{blue}{#1}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{A biophysical minimal model to investigate age-related changes in CA1 pyramidal cell electrical activity} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Erin C. McKiernan \textsuperscript{1,*},
Marco A. Herrera-Valdez \textsuperscript{2,**},
Diano F. Marrone \textsuperscript{3,4}
\\
\bigskip
\textbf{1} Departamento de F\'isica, Facultad de Ciencias, Universidad Nacional Aut\'onoma de M\'exico, Ciudad de México, CDMX, México
\\
\textbf{2} Laboratorio de Din\'amica, Biof\'isica y Fisiolog\'ia de Sistemas, Departamento de Matem\'aticas, Facultad de Ciencias, Universidad Nacional Aut\'onoma de M\'exico, Ciudad de México, CDMX, México
\\
\textbf{3} Department of Psychology, Wilfrid Laurier University, Waterloo, ON, Canada
\\
\textbf{4} McKnight Brain Institute, University of Arizona, Tucson, AZ, USA
\\
\bigskip

% Use the asterisk to denote corresponding authorship and provide email address in note below. 
% Comment from author ECM: We would like to include two corresponding authors, as below.
*emckiernan@ciencias.unam.mx (ECM) \\
**marcoh@ciencias.unam.mx (MAHV)

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Aging is a physiological process that is still poorly understood, especially with respect to effects on the brain. There are open questions about aging that are difficult to answer with an experimental approach. Underlying challenges include  the difficulty of recording \textit{in vivo} single cell and network activity simultaneously with submillisecond  resolution, and brain compensatory mechanisms triggered by genetic, pharmacologic, or behavioral manipulations. Mathematical modeling can help address some of these questions by allowing us to fix parameters that cannot be controlled experimentally and investigate neural activity under different conditions. We present a biophysical minimal model of CA1 pyramidal cells (PCs) based on general expressions for transmembrane ion transport derived from thermodynamical principles. The model allows directly varying the contribution of ion channels by changing their number. By analyzing the dynamics of the model, we find parameter ranges that reproduce the variability in electrical activity seen in PCs. In addition, increasing the L-type {\Ca} channel expression in the model reproduces age-related changes in \edit{electrical activity} that are qualitatively and quantitatively similar to those observed in PCs from aged animals. \edit{We also make predictions about age-related changes in PC bursting activity that, to our knowledge, have not been reported previously. We conclude that the model's biophysical nature, flexibility, and computational simplicity make it a potentially powerful complement to experimental studies of aging.}

% Author Summary not valid for PLOS ONE submissions.   
%\section*{Author summary}

@@

\section*{Introduction}
As we age, our brains undergo many changes
\cite{oh2010learning,rosenzweig2003impact}, but we understand relatively little about these and their effects on neural function. What does normal neurophysiological aging look like and what are the various stages? How does the electrical activity of neurons change and what are the biophysics underlying those changes?  How do aging neurons respond to input from other cells? Answering these questions is not just fundamental to understanding aging as a neurophysiological process, but also to understanding how this process may be altered in age-related disorders of clinical importance such as Alzheimer's \cite{fjell2014normal} and Parkinson's \cite{rodriguez2015parkinson} disease.

Many aging studies have focused on the hippocampus, an area of the brain involved in learning, memory formation, and spatial processing
\cite{oh2010learning,rosenzweig2003impact}. Aged rats \cite{barnes1985age,gage1984spatial,caprioli1991spatial} and humans \cite{wilkniss1997age} show impaired learning of hippocampal-dependent spatial tasks. Long-term potentiation (LTP), a proposed physiological substrate of memory formation, has been investigated in the hippocampus and its induction and maintenance shown to be impaired in aged rats
\cite{deupree1993age,rosenzweig1997role}. A short-term form of plasticity, frequency potentiation/facilitation (FP/FF), is also impaired in hippocampal pyramidal cells (PCs) from aged rats and correlates with learning deficits
\cite{landfield1978impaired}.

Plasticity changes and behavioral impairments may result in part from altered Ca$^{2+}$ signaling
in aged neurons
\cite{oh2010learning,rosenzweig2003impact}. Compared to CA1 PCs from young animals, PCs from aged animals show larger and longer post-burst afterhyperpolarizations (AHPs)
\cite{landfield1984prolonged,gant2009action,power2002age}. AHPs
are mediated by {\Ca}-dependent {\K} currents, which can act like brakes on the electrical activity of CA1 PCs
\cite{alger1980epileptiform,hotson1980calcium}. As a result, PCs show increased spike frequency adaptation and fire fewer action potentials (APs) in response to acute stimuli or during bursting activity 
\cite{gant2006early,moyer1992nimodipine,tombaugh2005slow}. Larger AHPs are associated with increased intracellular Ca$^{2+}$, mediated in part by {\Ca} entry via L-type channels
\cite{campbell1996aging,moyer1992nimodipine,power2002age,thibault2001elevated}. Aged
animals show increases in L-type channel expression and/or channel
density at the plasma membrane
\cite{herman1998up,thibault1996increase,veng2002regionally,nunez2014surface}.
Animals with higher {\Ca} channel density perform poorly in spatial
tasks  \cite{thibault1996increase}, while blockers of L-type channels can
restore learning and plasticity in older animals
\cite{norris1998reversal,sandin1990aging}.

It is not well understood how changes in ion channel gene expression and hippocampal PC excitability may affect neuron responsiveness and microcircuit output. In part, this is due
to challenges inherent in performing the needed experiments. Single PCs are difficult
to access in intact animals where hippocampal microcircuit function is
preserved. It is also difficult to tease apart the influence of the many different neurophysiological factors that change during aging. Mathematical modeling provides \edit{a} means to understand more about the effects of aging on hippocampal cellular excitability by controlling factors we cannot control experimentally.

Our previous work shows that mathematical expressions for different passive and active ion transport mechanisms can be derived from first principles of thermodynamics \cite{herrera2012membranes,herrera2013relating} using a common functional form \cite{herrera2018thermodynamic}. This results in a realistic representation of ionic flow across the membrane, and allows the model to reproduce phenomena such as rectification of ion currents seen in recordings. We present a model that reproduces the diversity of firing patterns in CA1 PC recordings, including adaptive firing, stimulus-induced bursting, and spontaneous bursting \cite{mckiernan2017ca1}. In addition, we reproduce several electrophysiological characteristics of aging by varying the expression of Ca$^{2+}$ channels in the model, and make predictions about bursting activity in aged CA1 PCs, which to our knowledge has not been reported. We believe this model is ideal to further study the effects of various biophysical changes in CA1 PCs during aging, as well as potentially forming the basis for biophysical, yet computationally inexpensive, network models.

\section*{Materials and methods}

\subsection*{Model}
To simulate the electrical activity of CA1 PCs, we used an extended version of a two-dimensional, biophysical model previously developed and characterized by two of the present authors  \cite{herrera2012membranes,herrera2013relating,herrera2018thermodynamic}. \edit{The present model differs from our previous formulations in that it includes \Ca dynamics, i.e. it is three-dimensional, allowing for additional behaviors like bursting \cite{avron1993basic}. In addition, it is specially tuned by incorporating specific channel variants and corresponding experimental data from hippocampal PCs, as described below.} The equations for the ionic currents are derived from first principles of thermodynamics. Previous modeling studies have shown that to reproduce firing behaviors such as spike frequency adaptation and bursting, the minimum number of variables is three \cite{hindmarsh1984model,avron1993basic}. 
In particular, {\Ca} dynamics are important for producing adaptation and burst firing in CA1 PCs (for review see \cite{mckiernan2017ca1}). 
The model dynamics are therefore described by three ordinary differential equations for the time-dependent changes in the transmembrane potential ($v$, in mV), the proportion of open {\K} channels ($w$ in [0,1]), and the intracellular {\Ca} concentration ($c$, in $\mu$M), respectively   \cite{herrera2018thermodynamic}. Based on a well-known relationship between voltage-dependent activation of delayed rectifier {\K} channels and inactivation of {\Na} channels, $w$ also represents the proportion of inactivated {\Na} channels \cite{rinzel1985excitation,avron1991minimal}.

It is assumed that the membrane potential changes due to currents produced by ions transported across the membrane. We take into account currents mediated by a voltage-gated inactivating {\Na} channel ($I_{NaT}$), \edit{voltage-gated L-type {\Ca} channel ($I_{CaL}$),} voltage- and {\Ca}-gated {\K} channels ($I_{DK}$ and $I_{SK}$, respectively), and a {\Na}/{\K}-ATPase ($I_{NaK}$). \edit{We also incorporate a forcing term ($I_F$), which can be used to stimulate the model PC and is explained in more detail below.} The time-dependent change in membrane potential can be written as
\begin{eqnarray}
\label{eq:dvdt}
C_m \partial_t v= I_{F} - I_{NaT}(v,w) - I_{CaL}(v,c) - I_{DK}(v,w) - I_{SK}(v,c) - I_{NaK}(v). 
\end{eqnarray}
with $\partial_t$ representing the instantaneous change with respect to time. $C_m$ (pF) is a constant representing the change in the density of charge around the membrane with respect to voltage, typically referred to as membrane capacitance in models based on electrical circuits  \cite{herrera2020nonequivalent}. Based on recordings from rat CA1 PCs, it is  assumed that $C_m=25$  pF \cite{groc2002vivo}.

All the currents in Eq.~\eqref{eq:dvdt} are modeled using the same generic functional form, a product
\begin{eqnarray}
    I_x = a_x G_x    \varphi_x, 
\end{eqnarray}
 $x \in \lrSet{NaT,CaL,DK,SK,NaK}$, where $a_x$ is a whole-cell current amplitude (pA), $G_x$ is a gating term (between 0~and~1), and $\varphi_x$ is an adimensional term describing the driving force for the transmembrane flux (Table~\ref{tab:ionFluxes}). The terms $a_x= s_x N_x$, $x \in \lrSet{NaT,CaL,DK,SK,NaK}$ are whole-cell current amplitudes with
$s_x$ (pA) representing the current flowing through a single channel (or pump), and $N_x$ representing the number of membrane proteins mediating the current (e.g. {number of {\K} channels}). Of interest, $s_x$ is $\sim$1 pA for most voltage-gated channels \cite{hille2001ion}, and is $\sim$5-10 pA for SK channels \cite{stocker20042+}.

Assuming that none of the currents in the model exhibit rectification  \cite{herrera2018thermodynamic}, which agrees with recordings of the included currents, the adimensional component of the transmembrane flux can  be simplified and written as
\begin{eqnarray}
    \varphi_x(v) = 2\eta_x \sinh\left(\eta_x \frac{v-v_x}{2v_T}\right),
\end{eqnarray} 
where $\eta_x$ represents the number of charges transported across the membrane in a single transport event, and $v_x$ is the reversal potential for the current,   
 $x \in \left\{NaT,CaL,DK,SK,NaK \right\}$. For channels, $v_x$ is the Nernst potential for the ion \cite{herrera2018thermodynamic}. Of note,  $\eta_x=1$ for $x \in \lrSet{DK,SK,NaK}$ and $\eta_{NaT}=-1$   \cite{herrera2018thermodynamic}, which gives 
\begin{eqnarray}
    \varphi_x(v) = 2 \sinh\left( \frac{v-v_{x}}{2v_T}\right),
\end{eqnarray} 
for $x \in \lrSet{NaT,DK,SK,NaK}$.
For {\Ca} channels, the total charge transported by one ion crossing the membrane from the extracelular space is $\eta_{CaL}=-2$, so
\begin{eqnarray}
    \varphi_{Ca}(v) = 4 \sinh\left( \frac{v-v_{Ca}}{v_T}\right).
\end{eqnarray}

The driving force for flux is assumed to be the same for DK and SK channels. Therefore, the label $K$ is used for both fluxes from here on.   
The thermal potential $v_T=kT/q$ (mV), where $k$ is Boltzmann's constant (mJ/$^o$K), $T$ is the  absolute temperature ($^o$K), and $q$ is the elementary charge  (Coulombs). The Boltzmann constant can be thought of as a scaling factor between macroscopic (thermodynamic temperature) and microscopic (thermal energy) physics \cite{kalinin2005boltzmann}. The reversal potentials for the different currents depend on the  Nernst potentials for each ion, as given by 
\begin{eqnarray}
v_x = \frac{v_T}{z_x} \textrm{ln} \left(\frac{[x]_o}{[x]_i} \right), \quad x \in  \left\{Na,Ca,K \right\}
\end{eqnarray}
where $z_x$ is the ion valence and $[x]_o$ and $[x]_i$ are the ion concentrations outside and inside the cell, respectively. The reversal potential for the {\Na}/{\K}-ATPase is given by ${v_{NaK}=v_{ATP}+3v_{Na}-2v_{K}}$ \cite{herrera2018thermodynamic}. The Nernst potentials for {\Na} and {\K} are assumed to be constant, but $v_{Ca}$ varies because the intracellular {\Ca} concentration is a state variable in the model.

%TRANSPORT TABLE
\begin{footnotesize}
\begin{table}[ht!]
\begin{adjustwidth}{-0.4in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
%\centering
\rowcolors{2}{white}{gray!10}
\midsepremove
\caption{\centering \textbf{Transport mechanisms included in the model}}
\begin{tabular}{l l c l l}
\toprule
\rowcolor{white}
Transport mechanism & Current & Amplitude ($a$)& Gating ($G$) & Flux $\varphi$ \\
\toprule
Transient {\Na} channels & $I_{NaT}(v,w)$ & $a_{Na}$ & $S_m(v) (1-w)$ & $\varphi_{Na}(v)$ \\
L-type {\Ca} channels & $I_{CaL}(v,c)$ & $a_{Ca}$ & $S_{n}(v)$ & $ \varphi_{Ca}(v)$  \\ 
Delayed rectifier {\K} channels& $I_{DK}(v,w)$ & $a_{DK}$ & $w$ & $\varphi_K(v)$  \\ 
SK {\Ca}-dependent {\K} channels& $I_{SK}(v,c)$ & $a_{SK}$& $ H_{SK}(c)$ & $ \varphi_K(v)$  
\\ 
{\Na}/{\K} pumps & $I_{NaK}(v)$ & $a_{NaK}$& 1 & $\varphi_{NaK}(v)$ \\ 
\bottomrule
\rowcolor{white}
\end{tabular}
\label{tab:ionFluxes}
%\vspace{0.2cm}
\begin{flushleft}
\footnotesize{All ion fluxes are given by a product of the form $I_x = a_x G_x \varphi_x$, where $a_x$, $G_x$, and $\varphi_x$ represent, respectively, the amplitude (normalized by membrane capacitance), gating, and driving force terms for the flux. $G_{NaK}=1$ represents saturation of the {\Na}/{\K} pumps. Note that inactivation of {\Na} channels is also represented by $w$ \cite{rinzel1985excitation,avron1991minimal}, so the proportion of non-inactivated {\Na} channels is $1-w$.}
\end{flushleft}
\end{adjustwidth}
\end{table}
\end{footnotesize}
